A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-10-23 imaging_demo/ms.tex
Remove usage of lambda in section 4
From a723a20 (master):
=====================

In ms.tex,

(*) Remove usage of lambda in section 4
Create subsections under section 4
From 9414f28 (master):
=====================

In ms.tex,

(*) Create subsections under section 4
fixed some typos and other small things.

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging
Correlator'' -- a highly parallelized Object Oriented Python package,
\footnote{EPIC package can be accessed at https://github.com/nithyanandan/EPIC}
now publicly available. Besides implementing the MOFF imaging algorithm it also
includes FX imaging using software holography technique and a simulator for
generating electric fields from a sky model.

Figure~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $E_1(f)\ldots E_\textrm{N}(f)$ ($\widetilde{E}_a$ in matrix
notation) for each of the antennas. Each of the complex antenna gains are
calibrated to correct the corresponding electric field spectra. These calibrated
electric fields are gridded using an antenna-based gridding convolution function
following which it is spatially Fourier transformed and squared to obtain images
for every time stamp. These images are then time-averaged to obtain the
accumulated image $I(f)$ ($I(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas which are then Fourier transformed by the F-engine to produce
    electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and
    gridded. The gridded electric fields $E_\textrm{g}(f)$ from each time series
    are imaged to produce an images $I_1(f)\ldots I_\textrm{N}(f)$. These images
    are time-averaged to obtain the final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Figure~\ref{fig:FX-flowchart} shows the flowchart for software holographic
imaging from a FX correlator. The antenna-based F-engine is identical to that in
the MOFF processing. The electric field spectra each antenna are then cross-
multiplied in the X-engine with those from all other antennas to obtain the
visibilities $V_\textrm{ij}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix notation).
They are calibrated and time-averaged to obtain $\langle V_\textrm{ij}(f)\rangle$
which are then gridded and imaged to obtain the image $I(f)$. The $I(f)$ obtained
from both techniques are identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of FX imaging in EPIC. The FX process flow shares the
    F-engine with the MOFF process. Following the F-engine, the electric fields
    pass through the X-engine to obtain visibilities $V_\textrm{ij}(f)$ which are
    calibrated and time-averaged. Then they are gridded to obtain the gridded
    visibilities $V_\textrm{g}(f)$ which are then imaged to obtain the image
    $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

The gridding kernel in the simplest case is given by the antenna aperture
illumination function, $\widetilde{B}(\mathbf{r}-\mathbf{r}_a)$, which can
specified either by a functional form or as a table of values against locations
around the antennas. A nearest neighbor mapping from all antenna footprints to
grid locations is created using an efficient k-d tree algorithm \citep{man99}.
There is no restriction here that the aperture illumination function has to be
identical across antennas.

In the most general case, this gridding kernel could contain information on the
$w$-projection effect, and even other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping has to be determined only once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{e})$ is described
as a transformation matrix from the space of measured electric fields
($\mathbf{e}$) to the antenna array grid denoted by $\mathbf{r}$. Since each
antenna occupies a footprint typically the size of its aperture,
$\mathbf{M}(\mathbf{r},\mathbf{e})$ which is generally of size
$N_\textrm{grid}\times N_\textrm{a}$, reduces to a sparse block-diagonal matrix
with only $N_\textrm{a}$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block. Figure~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

@@

\subsection{Calibration}
Calibration of direct imaging correlator remains an unresolved matter. Contrary
to the FX data flow, direct imagers mix the signals from all antennas before
averaging and writing to disk. It is therefore essential to apply gain solutions 
before the gridding step. Previous efforts have resorted to applying FX-generated
calibration solutions \citep{zhe14, fos14}, or integrate a dedicated FX correlator 
which periodically forms the full visibility matrix \citep{wij09, dev09}.

@@

The antenna array aperture illumination over the entire grid,
$\widetilde{W}(\mathbf{r})$, is obtained by a projection of the individual
antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{W}(\mathbf{r}) &= \sum_a \widetilde{W}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{e})\,\mathcal{I}(\mathbf{e}),
\end{align}
where, $\mathcal{I}(\mathbf{e})$ is a row of ones. This is achieved using
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process. Unless $\widetilde{W}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra. Thus,
\begin{align}
  \widetilde{E}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{e})\,E(\mathbf{e}),
\end{align}
where, $E(\mathbf{e})$ denotes the spectra of measured antenna electric fields.

In practice the grid is set to be a power of 2 such that the grid spacing is at
most $\lambda/2$ even at the highest frequency in the observing band.

@@

@@

@@

We study the effect of the differences in gridded weights on the image plane. 
The synthesized beam can be written as:
\begin{align}\label{eqn:syn-beam}
  B(\hat{\mathbf{s}}) &= \left| \sum_\mathbf{r} \widetilde{W}(\mathbf{r})\,e^{-i\frac{2\pi}{\lambda}(\mathbf{r}\cdot\hat{\mathbf{s}})}\right|^2, \\
  \textrm{or}, \quad B(l,m) &= \left| \sum_{r_x,r_y} \widetilde{W}(r_x,r_y)\,e^{-i\frac{2\pi}{\lambda}(r_x l + r_y m)}\right|^2 
\end{align}
where, $\hat{\mathbf{s}}\equiv (l,m)$, $\mathbf{r}\equiv (r_x,r_y)$ and 
$\widetilde{W}(\mathbf{r})$ is given by equation~\ref{eqn:gridding-convolution}.
If the errors in the determination of $r_x$ and $r_y$ are denoted by 
$\delta r_x$ and $\delta r_y$ respectively, the error on $B(l,m)$ can be 
estimated using a simple propagation of errors, where,
\begin{align}
  \frac{\partial{B}(l,m)}{\partial{r_x}} &= e^{-i\frac{2\pi}{\lambda}\mathbf{r}\cdot\hat{\mathbf{s}}}\left[\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i\frac{2\pi l}{\lambda}\widetilde{W}(\mathbf{r})\right] \label{eqn:syn-beam-pdev1}\\
  &\qquad\qquad \times \sum_{\mathbf{r}^\prime} \widetilde{W}^\star(\mathbf{r}^\prime)\,e^{+i\frac{2\pi}{\lambda}\mathbf{r}^\prime\cdot \hat{\mathbf{s}}} \\
  &= e^{-i\frac{2\pi}{\lambda}\mathbf{r}\cdot\hat{\mathbf{s}}}\left[\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i\frac{2\pi l}{\lambda}\widetilde{W}(\mathbf{r})\right]\,W^\star(\hat{\mathbf{s}}), \\
  \textrm{and},\frac{\partial{B}(l,m)}{\partial{r_y}} &= e^{-i\frac{2\pi}{\lambda}\mathbf{r}\cdot\hat{\mathbf{s}}}\left[\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i\frac{2\pi m}{\lambda}\widetilde{W}(\mathbf{r})\right]\,W^\star(\hat{\mathbf{s}}). \label{eqn:syn-beam-pdev2}
\end{align}
It must be noted that although we start with a squaring operation in 
equation~\ref{eqn:syn-beam}, in equations~\ref{eqn:syn-beam-pdev1} through 
\ref{eqn:syn-beam-pdev2}, the summations are performed only when 
$\mathbf{r}\ne \mathbf{r}^\prime$ and therefore do not include the 
auto-correlations of the antenna weights. The error expected in $B(l,m)$ is:
\begin{align}
  \left|\delta B(l,m)\right|^2 &= \sum_{r_x}\left|\frac{\partial{B}(l,m)}{\partial{r_x}}\right|^2 \delta r_x^2 + \sum_{r_y}\left|\frac{\partial{B}(l,m)}{\partial{r_y}}\right|^2 \delta r_y^2 \\
  &= \delta r_x^2 \sum_{r_x}\left|\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i\frac{2\pi l}{\lambda}\widetilde{W}(\mathbf{r})\right|^2\,\left|W(\hat{\mathbf{s}})\right|^2 \nonumber\\
  &\qquad + \delta r_y^2 \sum_{r_y}\left|\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_y}} - i\frac{2\pi m}{\lambda}\widetilde{W}(\mathbf{r})\right|^2\,\left|W(\hat{\mathbf{s}})\right|^2
\end{align}
For the current simulations, 
\begin{align}
  \delta r_x = \delta r_y &\simeq 1, \\
  \widetilde{W}(r_x,r_y) &= \{0, \textrm{constant}\}, \\
  \textrm{and},\qquad \frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} &= \frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_y}} = 0.
\end{align}
Hence,
\begin{align}
  \left|\delta B(l,m)\right|^2 &\simeq \left(\frac{2\pi}{\lambda}\right)^2\left(l^2+m^2\right)\,B(\hat{\mathbf{s}})\,\sum_\mathbf{r}\,\left|\widetilde{W}(\mathbf{r})\right|^2, \\
\textrm{or},\qquad \left|\delta B(l,m)\right| &\simeq \left(\frac{2\pi}{\lambda}\right)\sqrt{l^2+m^2}\,\left|W(\hat{\mathbf{s}})\right|\,\left(\sum_\mathbf{r}\,\left|\widetilde{W}(\mathbf{r})\right|^2\right)^{1/2}, \label{eqn:syn-beam-error}
\end{align}
where, $B(\hat{\mathbf{s}}) \equiv \left|W(\hat{\mathbf{s}})\right|^2$. Thus, 
the error on the synthesized beam is proportional to the antenna power pattern
and increases linearly outward in radial bins of direction cosines.

@@

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated.eps}
  \caption{Current and instruments planned for future in parameter space of
    number of complex multiplies and adds with MOFF and FX.}
  \label{fig:parameter-space-computations-instruments}
\end{figure}

>>>>

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging
Correlator'' -- a highly parallelized Object Oriented Python package,
\footnote{EPIC package can be accessed at https://github.com/nithyanandan/EPIC}
now publicly available. Besides implementing the MOFF imaging algorithm it also
includes FX imaging using the software holography technique and a simulator for
generating electric fields from a sky model.

Figure~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $E_1(f)\ldots E_\textrm{N}(f)$ ($\widetilde{E}_a$ in matrix
notation) for each of the antennas. Each of the complex antenna gains are
calibrated to correct the corresponding electric field spectra. These calibrated
electric fields are gridded using an antenna-based gridding convolution function
after which it is spatially Fourier transformed and squared to obtain images
for every time stamp. These images are then time-averaged to obtain the
accumulated image $I(f)$ ($I(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas which are then Fourier transformed by the F-engine to produce
    electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and
    gridded. The gridded electric fields $E_\textrm{g}(f)$ from each time series
    are imaged to produce an images $I_1(f)\ldots I_\textrm{N}(f)$. These images
    are time-averaged to obtain the final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Figure~\ref{fig:FX-flowchart} shows the flowchart for software holographic
imaging from a FX correlator. The antenna-based F-engine is identical to that in
the MOFF processing. The electric field spectra from each antenna are then cross-
multiplied in the X-engine with those from all other antennas to obtain the
visibilities $V_\textrm{ij}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix notation).
They are calibrated and time-averaged to obtain $\langle V_\textrm{ij}(f)\rangle$
which are then gridded and imaged to obtain the image $I(f)$. The $I(f)$ obtained
from both techniques are identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of FX imaging in EPIC. The FX process flow shares the
    F-engine with the MOFF process. Following the F-engine, the electric fields
    pass through the X-engine to obtain visibilities $V_\textrm{ij}(f)$ which are
    calibrated and time-averaged. Then they are gridded to obtain the gridded
    visibilities $V_\textrm{g}(f)$ which are then imaged to obtain the image
    $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

The gridding kernel in the simplest case is given by the antenna aperture
illumination function, $\widetilde{B}(\mathbf{r}-\mathbf{r}_a)$, which can
be specified either by a functional form or as a table of values against locations
around the antennas. A nearest neighbor mapping from all antenna footprints to
grid locations is created using an efficient k-d tree algorithm \citep{man99}.
There is no restriction here that the aperture illumination function has to be
identical across antennas.

In the most general case, this gridding kernel could contain information on the
$w$-projection effect, and even other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping must only be determined once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{e})$ is described
as a transformation matrix from the space of measured electric fields
($\mathbf{e}$) to the antenna array grid denoted by $\mathbf{r}$. Since each
antenna occupies a footprint typically the size of its aperture,
$\mathbf{M}(\mathbf{r},\mathbf{e})$, which is generally of size
$N_\textrm{grid}\times N_\textrm{a}$, reduces to a sparse block-diagonal matrix
with only $N_\textrm{a}$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block. Figure~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

@@

\subsection{Calibration}
Calibration of direct imaging correlators remains an unresolved matter. Contrary
to the FX data flow, direct imagers mix the signals from all antennas before
averaging and writing to disk. It is therefore essential to apply gain solutions 
before the gridding step. Previous efforts have resorted to applying FX-generated
calibration solutions \citep{zhe14, fos14}, or integrating a dedicated FX correlator 
which periodically forms the full visibility matrix \citep{wij09, dev09}.

@@

The antenna array aperture illumination over the entire grid,
$\widetilde{W}(\mathbf{r})$, is obtained by a projection of the individual
antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{W}(\mathbf{r}) &= \sum_a \widetilde{W}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{e})\,\mathcal{I}(\mathbf{e}),
\end{align}
where, $\mathcal{I}(\mathbf{e})$ is a row of ones. This is achieved using
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process. Unless $\widetilde{W}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra. Thus,
\begin{align}
  \widetilde{E}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{e})\,E(\mathbf{e}),
\end{align}
where $E(\mathbf{e})$ denotes the spectra of measured antenna electric fields.
% The sentence below was already stated earlier.
%In practice the grid is set to be a power of 2 such that the grid spacing is at
%most $\lambda/2$ even at the highest frequency in the observing band.

@@

\subsection{Simulations}\label{sec:sim}

@@

\subsection{Differences in outputs}\label{sec:diff}

@@

We study the effect of the differences in gridded weights on the image plane. 
The synthesized beam can be written as:
\begin{align}\label{eqn:syn-beam}
  B(\hat{\mathbf{s}}) &= \left| \sum_\mathbf{r} \widetilde{W}(\mathbf{r})\,e^{-i2\pi\mathbf{r}\cdot\hat{\mathbf{s}}}\right|^2, \\
  \textrm{or}, \quad B(l,m) &= \left| \sum_{r_x,r_y} \widetilde{W}(r_x,r_y)\,e^{-i2\pi(r_x l + r_y m)}\right|^2 
\end{align}
where, $\hat{\mathbf{s}}\equiv (l,m)$, $\mathbf{r}\equiv (r_x,r_y)$ and 
$\widetilde{W}(\mathbf{r})$ is given by equation~\ref{eqn:gridding-convolution}.
If the errors in the determination of $r_x$ and $r_y$ are denoted by 
$\delta r_x$ and $\delta r_y$ respectively, the error on $B(l,m)$ can be 
estimated using a simple propagation of errors, where,
\begin{align}
  \frac{\partial{B}(l,m)}{\partial{r_x}} &= e^{-i2\pi\mathbf{r}\cdot\hat{\mathbf{s}}}\left[\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i 2\pi l\widetilde{W}(\mathbf{r})\right] \label{eqn:syn-beam-pdev1}\\
  &\qquad\qquad \times \sum_{\mathbf{r}^\prime} \widetilde{W}^\star(\mathbf{r}^\prime)\,e^{+i2\pi\mathbf{r}^\prime\cdot \hat{\mathbf{s}}} \\
  &= e^{-i2\pi\mathbf{r}\cdot\hat{\mathbf{s}}}\left[\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i2\pi l\widetilde{W}(\mathbf{r})\right]\,W^\star(\hat{\mathbf{s}}), \\
  \textrm{and},\frac{\partial{B}(l,m)}{\partial{r_y}} &= e^{-i2\pi\mathbf{r}\cdot\hat{\mathbf{s}}}\left[\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i2\pi m\widetilde{W}(\mathbf{r})\right]\,W^\star(\hat{\mathbf{s}}). \label{eqn:syn-beam-pdev2}
\end{align}
It must be noted that although we start with a squaring operation in 
equation~\ref{eqn:syn-beam}, in equations~\ref{eqn:syn-beam-pdev1} through 
\ref{eqn:syn-beam-pdev2}, the summations are performed only when 
$\mathbf{r}\ne \mathbf{r}^\prime$ and therefore do not include the 
auto-correlations of the antenna weights. The error expected in $B(l,m)$ is:
\begin{align}
  \left|\delta B(l,m)\right|^2 &= \sum_{r_x}\left|\frac{\partial{B}(l,m)}{\partial{r_x}}\right|^2 \delta r_x^2 + \sum_{r_y}\left|\frac{\partial{B}(l,m)}{\partial{r_y}}\right|^2 \delta r_y^2 \\
  &= \delta r_x^2 \sum_{r_x}\left|\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} - i2\pi l\widetilde{W}(\mathbf{r})\right|^2\,\left|W(\hat{\mathbf{s}})\right|^2 \nonumber\\
  &\qquad + \delta r_y^2 \sum_{r_y}\left|\frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_y}} - i2\pi m\widetilde{W}(\mathbf{r})\right|^2\,\left|W(\hat{\mathbf{s}})\right|^2
\end{align}
For the current simulations, 
\begin{align}
  \delta r_x = \delta r_y &\simeq 1, \\
  \widetilde{W}(r_x,r_y) &= \{0, \textrm{constant}\}, \\
  \textrm{and},\qquad \frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_x}} &= \frac{\partial{\widetilde{W}(r_x,r_y)}}{\partial{r_y}} = 0.
\end{align}
Hence,
\begin{align}
  \left|\delta B(l,m)\right|^2 &\simeq 4\pi^2\left(l^2+m^2\right)\,B(\hat{\mathbf{s}})\,\sum_\mathbf{r}\,\left|\widetilde{W}(\mathbf{r})\right|^2, \\
\textrm{or},\qquad \left|\delta B(l,m)\right| &\simeq 2\pi\sqrt{l^2+m^2}\,\left|W(\hat{\mathbf{s}})\right|\,\left(\sum_\mathbf{r}\,\left|\widetilde{W}(\mathbf{r})\right|^2\right)^{1/2}, \label{eqn:syn-beam-error}
\end{align}
where, $B(\hat{\mathbf{s}}) \equiv \left|W(\hat{\mathbf{s}})\right|^2$. Thus, 
the error on the synthesized beam is proportional to the antenna power pattern
and increases linearly outward in radial bins of direction cosines.

@@

\subsection{Application to LWA data}\label{sec:LWA-data}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated.eps}
  \caption{Current and planned instruments in parameter space of
    number of complex multiplies and adds with MOFF and FX.}
  \label{fig:parameter-space-computations-instruments}
\end{figure}
